\chapter{Document History}

% 
% 
% Copyright 2002 Jonathan Bartlett
% 
% Permission is granted to copy, distribute and/or modify this
% document under the terms of the GNU Free Documentation License,
% Version 1.1 or any later version published by the Free Software
% Foundation; with no Invariant Sections, with no Front-Cover Texts,
% and with no Back-Cover Texts.  A copy of the license is included in fdl.xml
% 

\begin{itemize}\item 12/17/2002 - Version 0.5 - Initial posting of book under GNU FDL 
\item 07/18/2003 - Version 0.6 - Added ASCII appendix, finished the discussion of the CPU in the Memory chapter, reworked exercises into a new format, corrected several errors.  Thanks to Harald Korneliussen for the many suggestions and the ASCII table. 
\item 01/11/2004 - Version 0.7 - Added C translation appendix, added the beginnings of an appendix of x86 instructions, added the beginnings of a GDB appendix, finished out the files chapter, finished out the counting chapter, added a records chapter, created a source file of common linux definitions, corrected several errors, and lots of other fixes 
\item 01/22/2004 - Version 0.8 - Finished GDB appendix, mostly finished w/ appendix of x86 instructions, added section on planning programs, added lots of review questions, and got everything to a completed, initial draft state. 
\item 01/29/2004 - Version 0.9 - Lots of editting of all chapters.  Made code more consistent and made explanations clearer.  Added some illustrations. 
\item 01/31/2004 - Version 1.0 - Rewrote chapter 9.  Added full index.  Lots of minor corrections. 
\item 04/18/2004 - Version 1.1 - Lots of minor updates based on reader comments.  Made cleared distinction between dynamic and shared libraries. 
\end{itemize}

